### 2011-01-31 paper/learningMethods.tex
Many edits...

\edit{this might no longer flow...fix} There are a plethora of methods
available for the creation of a vector motion generating function
$g(t)$, and we've chosen two for comparison in this study.

The first method is to choose a family of \emph{parametrized
  functions}.  By fixing the parameters to a particular set of values,
we obtain a deterministic motion function over time.  We tried several
parametrizations on the robot and, upon obtaining reasonable early
success, settled on one particular parametrization.  We dubbed this
the \emph{SineModel5}, as it has five parameters and employs a sine
wave as its root pattern.

@@

@@

\begin{table}[h]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter    & Description               & Range \\
\hline
\hline
\amp         & Amplitude                 & [0, 400] \\
\hline
$\tau$       & Period                    & [.5, 8] \\
\hline
$m_O$        & Outer motor multiplier    & [-2, 2] \\
\hline
$m_F$        & Front motor multiplier    & [-1, 1] \\
\hline
$m_R$        & Right motor multiplier    & [-1, 1] \\
\hline
\end{tabular}
\caption{The parameters used for the \emph{SineModel5} motion model.}
\tablabel{parameters}
\end{center}
\end{table}

Intuitively, SineModel5 starts with 8 identically sine waves of
amplitude $\amp$ and period $\tau$, multiplies the waves for all inner
motors by multIO, multiplies the waves for all front motors by multFB,
and multiplies the waves for all left motors by multLR.  To obtain the
actual motor position commands, these waves are offset by
appropriate fixed constants so that the base position (when the sine
waves are at 0) is approximately a crouch (the position shown in
\figref{crouchingRobot}).  Finally, in this motion model, the ninth
"hip" motor is kept at a neutral position.  Thus, the commanded
position for all motors, as a vector function of time, is:

@@

\noindent Here the fixed constants that determine the crouched
position are $C_I$ for inner motors, $C_O$ for outer motors, and $C_C$
for the center hip motor.

\edit{make figure to go here (or another section), show drawing with
  inner/outer, l/r, and f/b motors.}

The nine subscripted functions correspond to the nine labeled motors
in \figref{robotDiagram}

% used is a function parametrized by five numbers, shown in \figref{foo}

@@

The eight methods used, and their methods for choosing the next
$\vec{\theta}$ are:

\edit{write this.  Old quadratot Method section below alsdkfj }

@@

\edit{JMC:Bullets=narrower column=wasted space. subsubsection instead?}

@@

\item \emph{Neuroevolution (Evolving Artificial Neural Networks with HyperNEAT)}:

HyperNEAT is an indirect encoding for evolving artificial neural
networks (ANNs) that is inspired by the way natural organisms
develop~\cite{stanley2009hypercube}. It evolves Compositional Pattern
Producing Networks (CPPNs)~\cite{stanley2007compositional}, each of
which is a genome that encodes an ANN
phenotype~\cite{stanley2009hypercube}. Each CPPN is itself a directed
graph, where each node is a mathematical function, such as sine or
Gaussian. The nature of these functions can facilitate the evolution
of properties such as symmetry (e.g., an absolute value or Gaussian
function) and repetition (e.g., a sine
function)~\cite{stanley2009hypercube, stanley2007compositional}. The
signal on each link in the CPPN is multiplied by that link's weight,
which can alter its effect.

A CPPN is queried once for each link in the ANN substrate to determine
that link's weight. The inputs to the CPPN are the Cartesian
coordinates of both the source (e.g., $x = 2$, $y = 4$) and target
(e.g., $x = 3$, $y = 5$) nodes of a link and a constant bias
value. The CPPN takes these five values as inputs and produces one or
two output values, depending on the substrate topology. If there is no
hidden layer in the substrate, the single output is the weight of the
link between a source node on the input layer and a target node on the
output layer. If there is a hidden layer, the first output value
determines the weight of the link between the associated input
(source) and hidden layer (target) nodes, and the second output value
determines the weight of the link between the associated hidden
(source) and output (target) layer nodes. All pairwise combinations of
source and target nodes are iteratively passed as inputs to a CPPN to
determine the weight of each substrate link.

\edit{following paragraph is figure caption for file named:
  hyperneatExplanation.png} HyperNEAT Produces ANNs from CPPNs. ANN
weights are specified as a function of the geometric coordinates of
the source node and the target node for each connection. The
coordinates of these nodes and a constant bias are iteratively passed
to the CPPN to determine each connection weight. The CPPN has two
output values, which specify the weights for each connection layer as
shown.

HyperNEAT is capable of exploiting the geometry of a problem: because
the link values between nodes in the final ANN substrate are a
function of the geometric positions of those nodes, HyperNEAT can
exploit such information when solving a
problem~\cite{stanley2009hypercube, clune2009sensitivity,
  clune2011performance}. In the case of quadruped locomotion, this
property helped Hyper-NEAT produce gaits with front-back, left-right,
and four-way symmetries~\cite{clune2009evolving,
  clune2011performance}.

The evolution of the population of CPPNs occurs according to the
principles of the NeuroEvolution of Augmenting Topologies (NEAT)
algorithm~\cite{stanley2002evolving}, which was originally designed to
evolve ANNs. NEAT can be fruitfully applied to CPPNs because of their
structural similarity to ANNs. For example, mutations can add a node,
and thus a function, to a CPPN graph, or change its link weights. The
NEAT algorithm is unique in three main
ways~\cite{stanley2002evolving}. Initially, it starts with small
genomes that encode simple networks and slowly complexifies them via
mutations that add nodes and links to the network, enabling the
algorithm to evolve the network topology in addition to its
weights. Secondly, NEAT has a fitness sharing mechanism that preserves
diversity in the system and gives time for new innovations to be tuned
by evolution before competing them against more adapted
rivals. Finally, NEAT tracks historical information to perform
intelligent crossover while avoiding the need for expensive
topological analysis. A full explanation of NEAT can be found
in~\cite{stanley2002evolving}].

The ANN configuration follows previous studies that evolved quadruped
gaits with HyperNEAT in simulation~\cite{clune2011performance,
  clune2009evolving}, but was adapted to accommodate differences in
the robot. Specifically, the ANNs consists of three $3 \times 4$
Cartesian grids of nodes forming input, hidden, and output
layers. Adjacent layers were completely connected, meaning that there
were $(3 \times 4)^2= 288$ links in each substrate. The inputs to the
substrate were the current angles of each of the 9 joints of the robot
(described below) and a sine and cosine wave (to facilitate the
production of periodic behaviors). The sine and cosine waves were the
following function of time (t) in milli-seconds:
$sin(t/12.0)\pi$. This function produces numbers in the range [$-\pi,
  \pi$], which were then JASON FILL ME IN TO DESCRIBE HOW JOINT
COMMANDS WERE SPECIFIED.

\edit{are we mentioning that we gave HyperNEAT the center joint here or earlier?}
\edit{jason: edit this section to reflect the post-processing...}

\edit{add the following paragraph as a figure caption to SpiderANN.jpg}

ANN Configuration for HyperNEAT Runs. The first two columns of each
row of the input layer receive information about a single leg (the
current angle of its two joints). The final column provides a sine and
cosine wave to enable periodic movements and the angle of the center
joint. Evolution determines the function of the hidden-layer
nodes. The nodes in the output layer specify new joint angles for each
respective joint. The unlabeled node in the input and output layer is
ignored.

\acmFig{explore_dim_1}{1}{Fitness mean and standard deviation
  vs. dimension 1.  The circle is a common point in
  \figref{explore_dim_1} through \figref{explore_dim_5}}

The population size for HyperNEAT was 9 and runs lasted 20
generations. These numbers are small, but were necesary given how much
time it took to conduct evolution directly on a real robot. The
remaining parameters were identical to Clune et
al. ~\cite{clune2011performance}.

  % Preliminary HyperNEAT runs were promising and resulted in several
  % interesting gaits.

  % Unfortunately, the gaits generated by HyperNEAT
  % also tended to stress the robot more than typical gaits had before,
  % and the servos would often overheat and malfunction, requiring
  % restarts.  We think these issues may be addressed by adding a small
  % layer between the HyperNEAT strategy and the robot that disallows
  % quickly shifting commanded positions, and we hope to be able to test
  % these methods further once this filter is in place.

@@

\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
  in simulation.  This simulation was used to tune the SVM strategy's
  parameters before trying it on the physical robot.  The strategy
  quickly approaches the maximum simulated fitness of 100.}

>>>>

There are a plethora of methods available for the creation of the
vector motion generating function $g(t)$ introduced in
\secref{problemDefinition}, and we've chosen two for comparison in
this study.  The first method we evaluated was to use a family of
parametrized gaits, described in this section and the next, and the
second method was to use an evolved non-parametric gait, detailed in
\secref{hyperNeatMethod}.

By a \emph{parametrized gait}, we mean a gait produced by a
parametrized function $g(t; \vec{\theta})$. Fixing the parameter
$\vec{\theta}$ yields a deterministic motion function over time.  We
tried several parametrizations on the robot and, upon obtaining
reasonable early success, settled on one particular parametrization.
We dubbed this the \emph{SineModel5}, as it has five parameters and
employs a sine wave as its root pattern.

@@

\edit{make sure this text flows into the table}

@@

\begin{table}[h!]
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Parameter    & Description               & Range \\
\hline
\hline
\amp         & Amplitude                 & [0, 400] \\
\hline
$\tau$       & Period                    & [.5, 8] \\
\hline
$m_O$        & Outer motor multiplier    & [-2, 2] \\
\hline
$m_F$        & Front motor multiplier    & [-1, 1] \\
\hline
$m_R$        & Right motor multiplier    & [-1, 1] \\
\hline
\end{tabular}
\caption{The parameters used for the \emph{SineModel5} motion model.}
\tablabel{parameters}
\end{center}
\end{table}

Intuitively, SineModel5 starts with 8 identically sine waves of
amplitude $\amp$ and period $\tau$, multiplies the waves for all inner
motors by $m_o$, multiplies the waves for all front motors by $m_F$,
and multiplies the waves for all left motors by $m_R$.  To obtain the
actual motor position commands, these waves are offset by
appropriate fixed constants so that the base position (when the sine
waves are at 0) is approximately a crouch (the position shown in
\figref{robot_close.jpg}).  Finally, in this motion model, the ninth
``hip'' motor is kept at a neutral position.  Thus, the commanded
position for all motors, as a vector function of time, is:

@@

\noindent Here the fixed constants that determine the crouched
position are $C_I$ for inner motors, $C_O$ for outer motors, and $C_C$
for the center hip motor.  The nine subscripted functions correspond
to the nine labeled motors in \figref{topdown.png}.

@@

The seven methods used, and their methods for choosing the next
$\vec{\theta}$ are:

@@

%\edit{JMC:Bullets=narrower column=wasted space. subsubsection instead?}

@@

@@

### 2011-01-31 paper/results.tex
Many edits...

\section{Results and Discussion}

\edit{proofread this.}
\subsection{Maximizing Strategies}
The maximizing strategies implemented include uniform random hill climbing, Gaussian random hill climbing, policy gradient descent, and Nelder-Mead simplex. As a group, these strategies produced results similar to gaits randomly generated using the SineModel5 model. None of these strategies consistently outperformed the others in the three trials; A different strategy generated the best gait in each trial. Uniform random hill climbing produced the best gait (11.37 body lengths/minute) in the first trial, with the next best performing strategy (Nelder-Mead simplex) resulting in a gait of 8.51 body lengths/minute. The best gait in the second trial was actually generated randomly (17.26 body-lengths/minute). Gaussian random hill climbing, policy gradient descent, and Nelder-Mead simplex all produced similar best gaits in the second trial, each around 14 body lengths/minute. In the third trial, Nelder-Mead simplex and Gaussian random hill climbing significantly outperformed the other maximizing strategies, resulting in gaits of 14.83 and 13.40 body lengths/minute respectively. Overall, Nelder-Mead simplex performed the best, producing an average best gait of 12.32 body lengths/minute. Gaussian random hill climbing followed, with an average best gait of 10.03 body lengths/minute.

\subsection{Predictive Strategies}
The predictive strategies used were linear regression and support vector regression. In the first trial, linear regression produced a gait (27.58 body lengths/minute) that was over three times faster than the best gait produced by a maximizing strategy (uniform random hill climbing -- 11.37 body lengths/minute). The average of the best gait for the three trials of linear regression was 14.01 body lengths/minute, which was better than the average best gait for the best performing maximizing strategy (Nelder-Mead simplex -- 12.32 body lengths/minute).

@@

\subsection{Evolved Neural Network Strategy}
The evolved neural network strategy used was HyperNEAT. HyperNEAT produced the fastest gaits by far. Linear regression produced a gait in the first trial (27.58 body lengths/minute) that was on par with the best gaits generated by HyperNEAT, by no other maximizing or predictive strategy produced gaits close to those of HyperNEAT. Linear regression was the best performing maximizing or predictive strategy, but it resulted in an average best gait of only 14.01 body lengths/minute. HyperNEAT, on the other hand, resulted in an average best gait of 29.26 body lengths/minute. One issue encountered with HyperNEAT was that it generated gaits that were rougher on the robot than gaits generated by the other strategies. The gaits sometimes caused the motor servos to overheat, which causes the motor to behave differently than expected. Since an overheated motor produces a different walk that expected, a test was performed after each run to make sure all motors were working correctly. If a motor was not, the distance walked would be halved to penalize walks that were harmful to the robot.

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & A       & B      & C      &  Average \\
\hline                                                               
\hline                                                               
Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
\hline                                                                 
Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
\hline                                                                 
Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
\hline                                                                 
Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
\hline                                                                 
Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
\hline                                                                 
Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
\hline                                                                 
Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
\hline                                                                
Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
\hline
\end{tabular}
\caption{The best gaits found for each starting vector and algorithm,
  in body lengths per minute.}
\tablabel{results_table}
\end{center}
\end{table}

@@

\begin{table}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & A       & B      & C      &  Average \\
\hline                                                               
\hline                                                               
Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
\hline                                                                 
Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
\hline                                                                 
Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
\hline                                                                 
Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
\hline                                                                 
Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
\hline                                                                 
Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
\hline                                                                 
Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
\hline                                                                
Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
\hline
\end{tabular}
\caption{The best gaits found for each starting vector and algorithm,
  in body lengths per minute.}
\tablabel{results_table}
\end{center}
\end{table}

@@

\subsection{Discussion}
% Is your hypothesis supported? What conclusions do the results support
% about the strengths and weaknesses of your method compared to other 
% methods? How can the results be explained in terms of the underlying 
% properties of the algorithm and/or the data.

>>>>

\section{Results and Discussion}
\seclabel{results}

\edit{proofread this.}

\subsection{Exploration of Parameter Space}

\subsection{Maximizing Strategies}

The maximizing strategies implemented include uniform random hill
climbing, Gaussian random hill climbing, policy gradient descent, and
Nelder-Mead simplex. As a group, these strategies produced results
similar to gaits randomly generated using the SineModel5 model. None
of these strategies consistently outperformed the others in the three
trials; A different strategy generated the best gait in each
trial. Uniform random hill climbing produced the best gait (11.37 body
lengths/minute) in the first trial, with the next best performing
strategy (Nelder-Mead simplex) resulting in a gait of 8.51 body
lengths/minute. The best gait in the second trial was actually
generated randomly (17.26 body-lengths/minute). Gaussian random hill
climbing, policy gradient descent, and Nelder-Mead simplex all
produced similar best gaits in the second trial, each around 14 body
lengths/minute. In the third trial, Nelder-Mead simplex and Gaussian
random hill climbing significantly outperformed the other maximizing
strategies, resulting in gaits of 14.83 and 13.40 body lengths/minute
respectively. Overall, Nelder-Mead simplex performed the best,
producing an average best gait of 12.32 body lengths/minute. Gaussian
random hill climbing followed, with an average best gait of 10.03 body
lengths/minute.

\subsection{Predictive Strategies}

The predictive strategies used were linear regression and support
vector regression. In the first trial, linear regression produced a
gait (27.58 body lengths/minute) that was over three times faster than
the best gait produced by a maximizing strategy (uniform random hill
climbing -- 11.37 body lengths/minute). The average of the best gait
for the three trials of linear regression was 14.01 body
lengths/minute, which was better than the average best gait for the
best performing maximizing strategy (Nelder-Mead simplex -- 12.32 body
lengths/minute).

@@

\edit{say something about tuning SVM and the sim figure}

\acmFig{svm_sim_results}{1}{Results for the SVM regression strategy
  in simulation.  This simulation was used to tune the SVM strategy's
  parameters before trying it on the physical robot.  The strategy
  quickly approaches the maximum simulated fitness of 100.}

\subsection{Evolved Neural Network Strategy}

The evolved neural network strategy used was HyperNEAT. HyperNEAT
produced the fastest gaits by far. Linear regression produced a gait
in the first trial (27.58 body lengths/minute) that was on par with
the best gaits generated by HyperNEAT, by no other maximizing or
predictive strategy produced gaits close to those of HyperNEAT. Linear
regression was the best performing maximizing or predictive strategy,
but it resulted in an average best gait of only 14.01 body
lengths/minute. HyperNEAT, on the other hand, resulted in an average
best gait of 29.26 body lengths/minute. One issue encountered with
HyperNEAT was that it generated gaits that were rougher on the robot
than gaits generated by the other strategies. The gaits sometimes
caused the motor servos to overheat, which causes the motor to behave
differently than expected. Since an overheated motor produces a
different walk that expected, a test was performed after each run to
make sure all motors were working correctly. If a motor was not, the
distance walked would be halved to penalize walks that were harmful to
the robot.

\edit{take this out?}

@@

\begin{table*}
\begin{center}
\begin{tabular}{|r|c|c|c||c|}
\hline
                                         & A       & B      & C      &  Average \\
\hline                                                               
\hline                                                               
Previous hand-coded gait                 & --      & --     & --     &  5.16 \\
\hline                                                                 
Random search                            & 6.04    & 17.26  & 4.90   &  9.40 \\
\hline                                                                 
Uniform Random Hill Climbing             & 11.37   & 9.44   & 2.69   &  7.83 \\
\hline                                                                 
Gaussian Random Hill Climbing            & 3.10    & 13.59  & 13.40  &  10.03 \\
\hline                                                                 
Policy Gradient Descent                  & 0.68    & 14.69  & 3.60   &  6.32 \\
\hline                                                                 
Nelder-Mead simplex                      & 8.51    & 13.62  & 14.83  &  12.32 \\
\hline                                                                 
Linear Regression                        & 27.58   & 12.51  & 1.95   &  14.01 \\
\hline                                                                
Evolutionary Neural Network (HyperNEAT)  & 24.27     & 36.44    & 27.07   & 29.26  \\
\hline
\end{tabular}
\caption{The best gaits found for each starting vector and algorithm,
  in body lengths per minute.}
\tablabel{results_table}
\end{center}
\end{table*}

@@

\subsection{Discussion}

% Is your hypothesis supported? What conclusions do the results support
% about the strengths and weaknesses of your method compared to other 
% methods? How can the results be explained in terms of the underlying 
% properties of the algorithm and/or the data.
